The Hyperedge Event Model

### 2018-09-05 paper/revision1/ba-sample.tex
Revision 1 folder created

	A major limitation of existing dynamic network models is that they apply to edges with one sender and one receiver. Dynamic network data often naturally arise as ``hyperedges'' \citep{karypis1999multilevel,ghoshal2009random,zlatic2009hypergraph,zhang2010hypergraph} that include one sender and multiple receivers or one receiver and multiple senders. For example, in email networks \citep{newman2002email}, each email encodes a hyperedge from one sender to one or more receviers. Networks formed between neurons via axons and dendrites involve hyperedges with one sender and multiple receivers (axons) or one receiver and multiple senders (dendrites) \citep{partzsch2012developing}. Networks formed through the cosponsorhip of legislative bills \citep{fowler2006legislative} involve hyperedges with multiple senders (cosponsors) and one receiver (sponsor). Economic sanctions between countries \citep{cranmer2014reciprocity} induce networks with hyperedges between multiple sending countries and one target country. Existing models require researchers to alter hyperedge data to fit with the pairwise edge structure of the model. For instance, \cite{PerryWolfe2012} treat multicast interactions---one type of directed hyperedge which involves one sender and one or more receivers---via duplication (i.e., obtain pairwise interactions from the original multicast), to construct an approximate likelihood function in their inferential framework for model parameters. Similarly, \cite{fan2009learning} duplicate emails sent from one sender to one or more receivers and randomly jitter the sent times in order to avoid violating the assumption that two events cannot occur at the exact same time. % There have been numerous studies for hyperedges under the hypergraph theory \citep{karypis1999multilevel} in phyics literature, including a random hypergraph model \citep{ghoshal2009random}, but most of them are designed specifically for engineering applications---e.g., multiple tagging networks \citep{zlatic2009hypergraph,zhang2010hypergraph}, cellular networks \citep{klamt2009hypergraphs}, and personalized recommendations \citep{zhang2010personalized,blattner2009b}---with lack of extensive efforts to incorportate hyperedge modeling into general statistical framework.

@@

	The hyperedge event model (HEM) specifies a generative process for $E$ unique hyperedge events that occur between $A$ nodes. A single hyperedge event is indexed by $e \in [E]$---where $[E]$ denotes a categorical set with $E$ levels $[E] = \{1,\ldots,E\}$---and consists of three components: the sender $s_e \in [A]$, an indicator vector of receivers $\boldsymbol{r}_e$---where $r_{ej}=1$ if $j \in [A]$ is a receiver of hyperedge event $e$ and 0 otherwise---and the timestamp $t_e \in (0, \infty)$. For simplicity, we assume that events are ordered by time such that $t_e \leq t_{e+1}$. While the model can be applied to two types of hyperedge events---events involving (1) one sender and one or more receivers, and (2) one or more senders and one receiver---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). One notable feature of our generative process is that we draw auxiliary variables that serve as candidate data. Data is generated from the HEM through a sampling process applied to the auxiliary variables. The auxiliary variables drawn for event $e$ include, for each possible sender $i \in [A]$, a time increment from event $e-1$ at which sender $i$ would create event $e$, and an $A-1$ length vector indicating which nodes would be the receivers of event $e$ if it were directed by sender $i$.  The data generated for event $e$ under the HEM corresponds to the sender that would create event $e$ the soonest---at the smallest time increment from event $e-1$. The receivers of event $e$ generated under the HEM correspond to those receivers toward which the sender with the minimum time increment would have directed event $e$. We explain these steps in more detail below. For hyperedge events that involve one receiver and one or more senders, we treat $s_e$ to be an indicator vector of senders $\boldsymbol{s}_e$ and $r_e$ to be the single receiver, and then follow the alternative generative process provided in Appendix A, which we derive as a simple reversal of the process used for multicasts (i.e., one sender and multiple receivers).~\looseness=-1

	\subsection{Candidate receivers}\label{subsec: Tie}
	% For every possible sender--receiver pair $(i,j)$ where $i \!\neq\! j$, we define the ``receiver intensity"---i.e., the conditional logit of the probability that hyperedge event $e$ is being directed from sender $i$ to receiver $j$---as a linear combination of statistics relevant to the receiver selection process:
	For every possible sender--receiver pair $(i,j)$ where $i \!\neq\! j$, we define the ``receiver intensity" as a linear combination of statistics relevant to the receiver selection process:
	\begin{equation}
		\lambda_{iej} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{iej},
	\end{equation}
	where $\boldsymbol{b}$ is a $P$-dimensional vector of coefficients and $\boldsymbol{x}_{iej}$ is a set of receiver selection features. As we show below, this intensity contributes to the probability that $i$ directs event $e$ towards receiver $j$. The features $\boldsymbol{x}_{iej}$ can capture common network processes like popularity, reciprocity, and transitivity, as well as the effects of attributes of the sender and receivers (e.g., their gender), or attributes of sender--receiver pairs (e.g., whether the sender is a supervisor of the receiver's). We also include a normally distributed intercept term to account for the average (or baseline) number of receivers: $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$.~\looseness=-1

	The HEM assumes that the sender of each hyperedge event is the sender that would initiate their respective event with the greatest urgency (i.e., the earliest timestamp). Our model thus assumes that for every event $e$, every possible sender $i$ generates a candidate receiver set that would be the receiver set of event $e$ if sender $i$ were the sender. For an event $e$, we first define an $A\times A$ matrix $\boldsymbol{u}_e$ where the $i^{\textrm{th}}$ row denotes sender $i$'s receiver vector of zeros and 1's---i.e., 1's indicate the nodes to which $i$ intends to direct event $e$. We then assume that each receiver vector $\boldsymbol{u}_{ie}$ comes from a modification of the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model that has been used to model graphs in which the state of each edge indicator is drawn independently from an edge-specific Bernoulli distribution. In order to avoid drawing hyperedge events with no receivers, we define a probability measure ``MB$_{G}$" motivated by the Gibbs measure \citep{fellows2017removing}. The probability measure we define amounts to a non-empty Gibbs measure, in which the all-zero vector is excluded from the support of the multivariate Bernoulli distribution. As a result, this measure helps us to (1) allow a sender to select multiple receivers for a single event, (2) force the sender to select at least one receiver, and (3) ensure a tractable normalizing constant for the receiver selection distribution. To be specific, we draw a binary vector $\boldsymbol{u}_{ie}= (u_{ie1},
	\ldots, u_{ieA})$ 
	\begin{equation} \boldsymbol{u}_{ie}  \sim
		\mbox{MB}_{G}(\boldsymbol{\lambda}_{ie}),
	\end{equation}
	where $\boldsymbol{\lambda}_{ie}= (\lambda_{ie1},
	\ldots, \lambda_{ieA})$. In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ie})$ as
	\begin{equation}
		\begin{aligned}
			&\Pr(\boldsymbol{u}_{ie}|\,\boldsymbol{b}, \boldsymbol{x}_{ie}) = \frac{1}{Z(\boldsymbol{\lambda}_{ie})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ie}\rVert_1 > 0 )\big) + \sum_{j\neq i} \lambda_{iej}u_{iej}\Big) ,
		\end{aligned}
		\label{eqn:Gibbs}
	\end{equation}
	where $Z(\boldsymbol{\lambda}_{ie})$ is the normalizing constant, $\lVert \cdot \rVert_1$ is the $\ell_1$-norm, and the log-indicator term $\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ie}\rVert_1 > 0 )\big)$ ensures that empty receiver sets are excluded from the distribution's support. These modeling assumptions facilitate efficient posterior inference since we can derive a closed form expression for the normalizing constant---i.e., $Z(\boldsymbol{\lambda}_{ie})= \prod_{j\neq i} \big(\mbox{exp}(\lambda_{iej}) + 1\big)-1$---and thus do not need to perform brute-force summation over the support of $\boldsymbol{u}_{ie} \in [0,1]^A$. We provide detailed derivation steps for the normalizing constant $Z(\boldsymbol{\lambda}_{ie})$ in Appendix B.

@@

	\subsection{Candidate timestamps}\label{subsec:Time}
	To generate timing for each event, the HEM first draws a candidate timestamp at which the event would be created given the candidate sender and receiver combinations. The timing rate for sender $i$ and event $e$ is
	\begin{equation}
		\mu_{ie} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ie}),
	\end{equation}
	where $\boldsymbol{\eta}$ is a $Q$-dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features---covariates that could affect timestamps of events, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.

	In modeling ``when," we do not directly model the timestamp $t_e$. Instead, we assume that each sender's ``time increment"---i.e., waiting time to next event since $t_{e-1}$---is drawn from a specific exponential family distribution. We define the time increment from event $e-1$ to event $e$ as $\tau_{e}$ (i.e., $\tau_{e}= t_e-t_{e-1}$) and specify the distribution of candidate timestamps with sender-specfic mean $\mu_{ie}$. Following the generalized linear model (GLM) framework \citep{nelder1972generalized}, we assume the mean and variance of the $\tau_{ie}$ satistify~\looseness=-1
	\begin{equation}
		\begin{aligned}
			E(\tau_{ie}) &= \mu_{ie},\\
			V(\tau_{ie}) &= V(\mu),
		\end{aligned}
	\end{equation}
		% \footnote{The log-normal distribution is not exponential family but can be used via modeling of $\log(\tau_e)$.} 
	where $\tau_{ie}$ here is a positive real number. Possible choices of distribution include exponential, Weibull, gamma, and log-normal distributions, which are commonly used in time-to-event modeling \citep{rao2000applied,rizopoulos2012joint}. Based on the specific distribution, we may need other latent variables to draw the time increment, to account for the variance of time increments, beyond the coefficients for the features used to model the rate. $V(\mu)$---e.g., the shape parameter $k$ for the Weibull, the shape parameter $\theta$ for the gamma, and the variance parameter $\sigma_\tau^2$ for the log-normal. We use $f_\tau(\cdot; \mu, V(\mu))$ and $F_\tau(\cdot; \mu, V(\mu))$ to denote the probability density function (p.d.f) and cumulative density function (c.d.f), respectively, with mean $\mu$ and variance $V(\mu)$.~\looseness=-1

	\subsection{Senders, receivers, and timestamps}\label{subsec:Observed}
	Finally, our model assumes that the observed sender, receivers, and timestamp of hyperedge event $e$ are generated by selecting the sender--receiver-set pair with the smallest time increment \citep{snijders1996stochastic}:
	\begin{equation}
		\begin{aligned}
			s_e &= \mbox{argmin}_{i}(\tau_{ie}),\\
			\boldsymbol{r}_e &= \boldsymbol{u}_{s_e e},\\
			t_e &=t_{e-1} + \tau_{s_e e}.
		\end{aligned}
	\end{equation}
	\begin{algorithm}[!t]
		\spacingset{1}
		\SetAlgoLined
		\caption{Generative process: one sender and one or more receivers}
		\begin{algorithmic}
			\STATE \textbf{Input}: number of events and nodes $(E, A)$, covariates $(\boldsymbol{x}, \boldsymbol{y})$, and coefficients $(\boldsymbol{b}, \boldsymbol{\eta})$
			\vskip 0.1in
			\FOR{$e=1$ to $E$}
			\FOR{$i=1$ to $A$}
			\FOR{$j=1$ to $A$ ($j \neq i$)}
			\STATE	set $\lambda_{iej} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{iej}$
			\ENDFOR
			\STATE	draw $\boldsymbol{u}_{ie}  \sim
			\mbox{MB}_G(\boldsymbol{\lambda}_{ie})$
			\STATE		set $\mu_{ie} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ie})$
			\STATE		draw $\tau_{ie} \sim f_\tau(\mu_{ie}, V(\mu))$
			\ENDFOR

			\IF{$n \geq 2$ tied events} 
			\STATE	set $s_e,\ldots, s_{e+n-1}=\mbox{argmin}_{i}(\tau_{ie}),$
			\STATE	set $\boldsymbol{r}_e=\boldsymbol{u}_{s_e e},\ldots,\boldsymbol{r}_{e+n-1}=\boldsymbol{u}_{s_{e+n-1} e}$
			\STATE	set $t_e, \ldots, t_{e+n-1}=t_{e-1} + \min_i\tau_{ie}$
			\STATE		jump to $e = e+n$
			\ELSE
			\STATE	set $s_e= \mbox{argmin}_{i}(\tau_{ie})$
			\STATE		set $\boldsymbol{r}_e = \boldsymbol{u}_{s_e e}$
			\STATE	set $t_e =t_{e-1} + \min_i\tau_{ie}$
			\ENDIF
			\ENDFOR
		\end{algorithmic}
		\label{alg:generative}
	\end{algorithm}
		Therefore, the HEM assumes a sender-driven process---i.e., the receivers and timestamp of an event are jointly determined by the sender's urgency to direct the event to those selected receivers. Note that our generative process allows for tied events. In the case of tied events---i.e., multiple senders draw exactly the same candidate timestamps---we assume that all events are generated and occur simultaneously. Algorithm \ref{alg:generative} summarizes the entire generative process for hyperedge events with one sender and one or more receivers, and Figure \ref{figure:diagram} presents an illustrative example on how the $e^{\textrm{th}}$ event is generated, assuming $t_{e-1} = 0$ and $A=5$.~\looseness=-1	
		\begin{figure}[!t]
			\centering
			\includegraphics[width=0.625\textwidth]{img/diagramnew.png}	
			\caption {An illustrative example of the generative process of the HEM.}
			\label{figure:diagram}
		\end{figure}

	\section{Posterior inference}\label{sec:inference}
	In this section we describe how we invert the generative process to obtain the posterior distribution over the latent variables---candidate receivers $\{\boldsymbol{u}_e\}_{e=1}^E$, coefficients for receiver selection features $\boldsymbol{b}$, and coefficients for event timing features $\boldsymbol{\eta}$---conditioned on the observed data $\{(s_e, \boldsymbol{r}_e, t_e)\}_{e=1}^E$, covariates $\{(\boldsymbol{x}_e, \boldsymbol{y}_e)\}_{e=1}^E$, and hyperparamters $(\boldsymbol{\mu}_b, \Sigma_b, \boldsymbol{\mu}_\eta, \Sigma_\eta)$. We draw the samples using Markov chain Monte Carlo (MCMC) methods, repeatedly resampling the value of each latent variable from its conditional posterior via a Metropolis-within-Gibbs sampling algorithm. In the next subsection, we provide each latent variable's conditional posterior along with pseudocode of MCMC in Algorithm \ref{alg:MCMC}. We also evaluate the correctness of both our mathematical derivations and software implemenation using the prior--posterior simulator test of \cite{geweke2004getting}.

	\subsection{Conditional posteriors}\label{subsec:conditionaldist}

@@

	Unlike the candidate receivers above, the conditional posterior for $\boldsymbol{b}$ does not have a closed form; however $\boldsymbol{b}$ may instead be re-sampled using the Metropolis--Hastings (MH) algorithm. Assuming an uninformative prior (i.e., $N({0},\infty)$), the conditional posterior for $\boldsymbol{b}$ is proportional to~\looseness=-1

@@

	Likewise, we use the MH algorithm to update the latent variable $\boldsymbol{\eta}$. Assuming an uninformative prior $\boldsymbol{\eta}$ (i.e., $N({0},\infty)$), the conditional posterior for an untied event case is proportional to~\looseness=-1

@@

	\subsection{Getting it Right (GiR) test} \label{subec:GiR}
	Software development is integral to the objective of applying our model to real world data. Code review is a valuable process in any research computing context, and the prevalence of software bugs in statistical software is well documented \citep[e.g., ][]{altman2004numerical,mccullough2009accuracy}.  With highly complex models such as the HEM, there are many ways in which software bugs can be introduced and go unnoticed. As such, we present a joint analysis of the integrity of our generative model, sampling equations, and software implementation.

>>>>

	A major limitation of existing dynamic network models is that they apply to edges with one sender and one receiver. Dynamic network data often naturally arise as ``hyperedges'' \citep{karypis1999multilevel,ghoshal2009random,zlatic2009hypergraph,zhang2010hypergraph} that include one sender and multiple receivers or one receiver and multiple senders. For example, in email networks \citep{newman2002email}, each email encodes a hyperedge from one sender to one or more receivers. Networks formed between neurons via axons and dendrites involve hyperedges with one sender and multiple receivers (axons) or one receiver and multiple senders (dendrites) \citep{partzsch2012developing}. Networks formed through the cosponsorship of legislative bills \citep{fowler2006legislative} involve hyperedges with multiple senders (cosponsors) and one receiver (sponsor). Economic sanctions between countries \citep{cranmer2014reciprocity} induce networks with hyperedges between multiple sending countries and one target country. Existing models require researchers to alter hyperedge data to fit with the pairwise edge structure of the model. For instance, \cite{PerryWolfe2012} treat multicast interactions---one type of directed hyperedge which involves one sender and one or more receivers---via duplication (i.e., obtain pairwise interactions from the original multicast), to construct an approximate likelihood function in their inferential framework for model parameters. Similarly, \cite{fan2009learning} duplicate emails sent from one sender to one or more receivers and randomly jitter the sent times in order to avoid violating the assumption that two events cannot occur at the exact same time. % There have been numerous studies for hyperedges under the hypergraph theory \citep{karypis1999multilevel} in phyics literature, including a random hypergraph model \citep{ghoshal2009random}, but most of them are designed specifically for engineering applications---e.g., multiple tagging networks \citep{zlatic2009hypergraph,zhang2010hypergraph}, cellular networks \citep{klamt2009hypergraphs}, and personalized recommendations \citep{zhang2010personalized,blattner2009b}---with lack of extensive efforts to incorportate hyperedge modeling into general statistical framework.

@@

	The hyperedge event model (HEM) specifies a generative process for $E$ unique hyperedge events that occur between $A$ nodes. A single hyperedge event is indexed by $e \in [E]$---where $[E]$ denotes a categorical set with $E$ levels $[E] = \{1,\ldots,E\}$---and consists of three components: the sender $s_e \in [A]$, an indicator vector of receivers $\boldsymbol{r}_e$---where $r_{ej}=1$ if $j \in [A]$ is a receiver of hyperedge event $e$ and 0 otherwise---and the timestamp $t_e \in (0, \infty)$. For simplicity, we assume that events are ordered by time such that $t_e \leq t_{e+1}$. While the model can be applied to two types of hyperedge events---events involving (1) one sender and one or more receivers, and (2) one or more senders and one receiver---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). One notable feature of our generative process is that we draw auxiliary variables that serve as candidate data. Data is generated from the HEM through a sampling process applied to the auxiliary variables. The auxiliary variables drawn for event $e$ include, for each possible sender $i \in [A]$, a time increment from event $e-1$ at which sender $i$ would create event $e$, and an $A-1$ length vector indicating which nodes would be the receivers of event $e$ if it were directed by sender $i$.  The data generated for event $e$ under the HEM corresponds to the sender that would create event $e$ the soonest---at the smallest time increment from event $e-1$. The receivers of event $e$ generated under the HEM correspond to those receivers toward which the sender with the minimum time increment would have directed event $e$. We explain these steps in more detail below. Figure \ref{figure:diagram} presents an illustrative example on how the $e^{\textrm{th}}$ event is generated, assuming $t_{e-1} = 0$ and $A=5$. For hyperedge events that involve one receiver and one or more senders, we treat $s_e$ to be an indicator vector of senders $\boldsymbol{s}_e$ and $r_e$ to be the single receiver, and then follow the alternative generative process provided in Appendix A, which we derive as a simple reversal of the process used for multicasts (i.e., one sender and multiple receivers).~\looseness=-1 
				\includegraphics[width=0.625\textwidth]{img/diagramnew.png}	
				\caption {An illustrative example of the generative process of the HEM.}
				\label{figure:diagram}
	\subsection{Candidate receivers}\label{subsec: Tie}
	% For every possible sender--receiver pair $(i,j)$ where $i \!\neq\! j$, we define the ``receiver intensity"---i.e., the conditional logit of the probability that hyperedge event $e$ is being directed from sender $i$ to receiver $j$---as a linear combination of statistics relevant to the receiver selection process:
	The HEM assumes that the sender of each hyperedge event is the sender that would initiate their respective event with the greatest urgency (i.e., the earliest timestamp). Our model thus assumes that for every event $e$, every possible sender $i$ generates a candidate receiver set that would be the receiver set of event $e$ if sender $i$ were the sender. For an event $e$, we first define an $A\times A$ matrix $\boldsymbol{u}_e$ where the $i^{\textrm{th}}$ row denotes sender $i$'s receiver vector of zeros and 1's---i.e., 1's indicate the nodes to which $i$ intends to direct event $e$. We then assume that each receiver vector $\boldsymbol{u}_{ie}$ comes from a modification of the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model that has been used to model graphs in which the state of each edge indicator is drawn independently from an edge-specific Bernoulli distribution. Motivated by the Gibbs measure \citep{fellows2017removing}, we define a probability measure ``MB$_{G}$" in order to avoid drawing hyperedge events with no receiver. This probability measure amounts to a non-empty Gibbs measure, in which the all-zero vector is excluded from the support of the multivariate Bernoulli distribution. As a result, this measure helps us to (1) allow a sender to select multiple receivers for a single event, (2) force the sender to select at least one receiver, and (3) ensure a tractable normalizing constant for the receiver selection distribution. To be specific, we draw a binary vector $\boldsymbol{u}_{ie}= (u_{ie1},
	\ldots, u_{ieA})$ for $i$th row of matrix $\boldsymbol{u}$ using
	\begin{equation} \boldsymbol{u}_{ie}  \sim
		\mbox{MB}_{G}(\boldsymbol{\lambda}_{ie}),
	where $\boldsymbol{\lambda}_{ie}= (\lambda_{ie1},
	\ldots, \lambda_{ieA})$ and $\lambda_{iej}$ is the linear combination of statistics relevant to the receiver selection process for every possible sender--receiver pair $(i,j)$ where $i \!\neq\! j$. In other words, we define
	\lambda_{iej} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{iej},
	where $\boldsymbol{b}$ is a $P$-dimensional vector of coefficients and $\boldsymbol{x}_{iej}$ is a set of receiver selection features. As we show below, $\lambda_{iej}$ contributes to the probability that $i$ directs event $e$ towards receiver $j$. The features $\boldsymbol{x}_{iej}$ can capture common network processes like popularity, reciprocity, and transitivity, as well as the effects of attributes of the sender and receivers (e.g., their gender), or attributes of sender--receiver pairs (e.g., whether the sender is a supervisor of the receiver's). We also include a normally distributed intercept term to account for the average (or baseline) number of receivers: $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$. ~\looseness=-1

	In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ie})$ as
	\begin{equation}
		\begin{aligned}
			&\Pr(\boldsymbol{u}_{ie}|\,\boldsymbol{b}, \boldsymbol{x}_{ie}) = \frac{1}{Z(\boldsymbol{\lambda}_{ie})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ie}\rVert_1 > 0 )\big) + \sum_{j\neq i} \lambda_{iej}u_{iej}\Big) ,
		\end{aligned}
		\label{eqn:Gibbs}
	\end{equation}
	where $Z(\boldsymbol{\lambda}_{ie})$ is the normalizing constant, $\lVert \cdot \rVert_1$ is the $\ell_1$-norm, and the log-indicator term $\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ie}\rVert_1 > 0 )\big)$ ensures that empty receiver sets are excluded from the distribution's support. These modeling assumptions facilitate efficient posterior inference since we can derive a closed form expression for the normalizing constant---i.e., $Z(\boldsymbol{\lambda}_{ie})= \prod_{j\neq i} \big(\mbox{exp}(\lambda_{iej}) + 1\big)-1$---and thus do not need to perform brute-force summation over the support of $\boldsymbol{u}_{ie} \in \{0,1\}^A$. We provide detailed derivation steps for the normalizing constant $Z(\boldsymbol{\lambda}_{ie})$ in Appendix B.

@@

	\subsection{Candidate timestamps}\label{subsec:Time}	
	 In modeling ``when," we do not directly model the timestamp $t_e$. Instead, we assume that each sender's ``time increment"---i.e., waiting time to next event since $t_{e-1}$---is drawn from a specific exponential family distribution. We assume that every sender $i \in [A]$ generates his/her own time increment from event $e-1$ to event $e$, which we define as $\tau_{ie}$. We further specify that the distribution of $\tau_{ie}$ has sender-specific mean $\mu_{ie}$ and common variance $V(\mu)$. Following the generalized linear model (GLM) framework \citep{nelder1972generalized}, we assume that sender $i$'s time increment $\tau_{ie}$ satisfy~\looseness=-1
	\begin{equation}
		\begin{aligned}
			E(\tau_{ie}) &= \mu_{ie},\\
			V(\tau_{ie}) &= V(\mu),
		\end{aligned}
	\end{equation}
		% \footnote{The log-normal distribution is not exponential family but can be used via modeling of $\log(\tau_e)$.} 
	where $\tau_{ie}$ here is a positive real number. Possible choices of distribution include exponential, Weibull, gamma, and log-normal distributions, which are commonly used in time-to-event modeling \citep{rao2000applied,rizopoulos2012joint}. Based on the specific distribution, we may need other latent variables to draw the time increment, to account for the variance of time increments $V(\mu)$---e.g., the shape parameter $k$ for the Weibull, the shape parameter $\theta$ for the gamma, and the variance parameter $\sigma_\tau^2$ for the log-normal---beyond the coefficients for the features used to model the rate. We use $f_\tau(\cdot; \mu, V(\mu))$ and $F_\tau(\cdot; \mu, V(\mu))$ to denote the probability density function (p.d.f) and cumulative density function (c.d.f), respectively, with mean $\mu$ and variance $V(\mu)$. Similar to Section \ref{subsec: Tie}, the timing rate $\mu_{ie}$ for sender $i$ and event $e$ follows
	\begin{equation}
	\mu_{ie} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ie}),
	\end{equation}
	where $\boldsymbol{\eta}$ is a $Q$-dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features---covariates that could affect timestamps of events, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.

	\subsection{Senders, receivers, and timestamps}\label{subsec:Observed}
		After the HEM draws the candidate timestamp vector $\boldsymbol{\tau}_e=\{\tau_{1e},\ldots,\tau_{Ae}\}$ at which the event would be created given the candidate sender and receiver combinations, finally our model assumes that the observed sender, receivers, and timestamp of hyperedge event $e$ are generated by selecting the sender--receiver-set pair with the smallest time increment:
	\begin{equation}
		\begin{aligned}
			s_e &= \mbox{argmin}_{i}(\tau_{ie}),\\
			\boldsymbol{r}_e &= \boldsymbol{u}_{s_e e},\\
			t_e &=t_{e-1} + \tau_{s_e e}.
		\end{aligned}
	\end{equation}
	\begin{algorithm}[!t]
		\spacingset{1}
		\SetAlgoLined
		\caption{Generative process: one sender and one or more receivers}
		\begin{algorithmic}
			\STATE \textbf{Input}: number of events and nodes $(E, A)$, covariates $(\boldsymbol{x}, \boldsymbol{y})$, and coefficients $(\boldsymbol{b}, \boldsymbol{\eta})$
			\vskip 0.1in
			\FOR{$e=1$ to $E$}
			\FOR{$i=1$ to $A$}
			\FOR{$j=1$ to $A$ ($j \neq i$)}
			\STATE	set $\lambda_{iej} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{iej}$
			\ENDFOR
			\STATE	draw $\boldsymbol{u}_{ie}  \sim
			\mbox{MB}_G(\boldsymbol{\lambda}_{ie})$
			\STATE		set $\mu_{ie} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ie})$
			\STATE		draw $\tau_{ie} \sim f_\tau(\mu_{ie}, V(\mu))$
			\ENDFOR

			\IF{$n \geq 2$ tied events} 
			\STATE	set $s_e,\ldots, s_{e+n-1}=\mbox{argmin}_{i}(\tau_{ie}),$
			\STATE	set $\boldsymbol{r}_e=\boldsymbol{u}_{s_e e},\ldots,\boldsymbol{r}_{e+n-1}=\boldsymbol{u}_{s_{e+n-1} e}$
			\STATE	set $t_e, \ldots, t_{e+n-1}=t_{e-1} + \min_i\tau_{ie}$
			\STATE		jump to $e = e+n$
			\ELSE
			\STATE	set $s_e= \mbox{argmin}_{i}(\tau_{ie})$
			\STATE		set $\boldsymbol{r}_e = \boldsymbol{u}_{s_e e}$
			\STATE	set $t_e =t_{e-1} + \min_i\tau_{ie}$
			\ENDIF
			\ENDFOR
		\end{algorithmic}
		\label{alg:generative}
	\end{algorithm}
		Therefore, the HEM assumes a sender-driven process---i.e., the receivers and timestamp of an event are jointly determined by the sender's urgency to direct the event to those selected receivers---which is similar to \cite{snijders1996stochastic} in that each actor gets his/her own rate function to determine ``who will make the next change in network." Note that our generative process allows for tied events. In the case of tied events---i.e., multiple senders draw exactly the same candidate timestamps---we assume that all events are generated and occur simultaneously. Algorithm \ref{alg:generative} summarizes the entire generative process for hyperedge events with one sender and one or more receivers.~\looseness=-1

	\section{Posterior inference}\label{sec:inference}
	In this section we describe how we invert the generative process to obtain the posterior distribution over the latent variables---candidate receivers $\{\boldsymbol{u}_e\}_{e=1}^E$, coefficients for receiver selection features $\boldsymbol{b}$, and coefficients for event timing features $\boldsymbol{\eta}$---conditioned on the observed data $\{(s_e, \boldsymbol{r}_e, t_e)\}_{e=1}^E$, covariates $\{(\boldsymbol{x}_e, \boldsymbol{y}_e)\}_{e=1}^E$, and hyperparameters $(\boldsymbol{\mu}_b, \Sigma_b, \boldsymbol{\mu}_\eta, \Sigma_\eta)$. We draw the samples using Markov chain Monte Carlo (MCMC) methods, repeatedly resampling the value of each latent variable from its conditional posterior via a Metropolis-within-Gibbs sampling algorithm. In the next subsection, we provide each latent variable's conditional posterior along with pseudocode of MCMC in Algorithm \ref{alg:MCMC}. We also evaluate the correctness of both our mathematical derivations and software implemenation using the prior--posterior simulator test of \cite{geweke2004getting} in Appendix C.

	%\subsection{Conditional posteriors}\label{subsec:conditionaldist}

@@

	Unlike the candidate receivers above, the conditional posterior for $\boldsymbol{b}$ does not have a closed form; however $\boldsymbol{b}$ may instead be re-sampled using the Metropolis--Hastings (MH) algorithm using the proposal distribution $N(\boldsymbol{0}, \delta_b\times I_P)$, where $\delta_b$ is the proposal variance parameter for $\boldsymbol{b}$. Assuming an uninformative prior (i.e., $N(\boldsymbol{0},\infty)$), the conditional posterior for $\boldsymbol{b}$ is proportional to~\looseness=-1

@@

	Likewise, we use the MH algorithm to update the latent variable $\boldsymbol{\eta}$ using the proposal distribution $N(\boldsymbol{0}, \delta_\eta\times I_Q)$, where $\delta_\eta$ is the proposal variance parameter for $\boldsymbol{\eta}$. Assuming an uninformative prior $\boldsymbol{\eta}$ (i.e., $N(\boldsymbol{0},\infty)$), the conditional posterior for an untied event case is proportional to~\looseness=-1

@@

	We evaluated the models' ability to predict out-of-sample events and timestamps on Montgomery County email data. We generated a train--test split of the data by randomly selecting 10\% of senders, receivers, and timestamp variables to be held out. Our model then imputed these missing variables during inference, sampling them from their conditional posterior along with the other latent variables. Algorithm \ref{alg:PPE} outlines this procedure in detail. We compare the predictive performance of two versions of our model, each with a different timing distribution over the time increments, by repeating the predictive assessments$N=500$ times where we make new predictions each time for the fixed held out data. We summarize the results of prediction experiments for missing senders, receivers, and timestamps in Figure \ref{figure:PPEresults}. First, we compare the posterior probability of correct senders for each of the missing emails $\{e:s_e=\mbox{NA}\}$, which corresponds to $\pi_{es_e}$ in Algorithm \ref{alg:PPE}. We call this measure the ``correct sender posterior probability." In Figure \ref{subfigure:PPEresults1}, we display boxplots for the distribution of mean correct sender posterior probability---i.e., $\hat{\pi}_{es_{e}} = \frac{1}{N}\sum_{n=1}^N \pi^{(n)}_{es_{e}}$---across the missing emails. The results show that both log-normal and exponential distributions achieve better predictive performance for missing senders compared to what is expected under random guess (i.e., choose one out of $A$ possible senders $=1/18$), with the log-normal model showing better performance than the exponential model. Secondly, since the receiver vector is binary, we compute $F_1$ scores for missing receiver indicators (i.e., all $e$ and $j$ with $r_{ej}$=NA) by taking the harmonic mean of precision and recall:
with TP denoting true positive (i.e., $\boldsymbol{r}^{\textrm{(obs)}}_{ej}=\boldsymbol{r}^{\textrm{(pred)}}_{ej}=1$), FN denoting false negative (i.e., $\boldsymbol{r}^{\textrm{(obs)}}_{ej}=1$ but $\boldsymbol{r}^{\textrm{(pred)}}_{ej}=0$), and FP denoting false positive (i.e., $\boldsymbol{r}^{\textrm{(obs)}}_{ej}=0$ but $\boldsymbol{r}^{\textrm{(pred)}}_{ej}=1$). Although the generative process for events (Section \ref{subsec: Tie}) is not directly affected by the choice of timestamp distribution, Figure \ref{subfigure:PPEresults2} reveals slight difference between log-normal and exponential in their performance in predicting missing receiver indicators, where log-normal on average outperforms exponential. Finally, the prediction error for the $d^{\textrm{th}}$ missing timestamp is estimated using the median of the absolute relative errors, often referred to as median absolute percentage error (MdAPE), across $N=500$ predictions:
